# Seasonal flow timeseries model<br> from paired flow and meterological/weather record
## 
## ALR July 2014<br>Conte Anadromous Fish Research Center
##   

### Load libraries, functions, etc
```{r load libraries and knitr settings, warning=FALSE, results='hide', message=FALSE}
library(lme4)
library(devtools)
library(knitr)
library(ggplot2)
library(plyr)

opts_chunk$set(comment=NA)
```


```{r load functions and save directory locations, warning=FALSE, results='hide', message=FALSE}
#utility functions, load (source) from my saved gist
source_gist("https://gist.github.com/anarosner/ba285306fc0ce9d812a5", sha1="b25a1b73e02cc2b2d2c590f6c0b2c9c9945fa980")

model_dir<-"C:/ALR/Models/MetToFlow"
model_data_dir<-"C:/ALR/Models_processed_data"

#load (source) this project's general functions
setwd(file.path(model_dir,"A_get_data"))  
purl(input="0_general_functions.Rmd",output="0_general_functions.R",documentation=0)
source("0_general_functions.R")
file.remove("0_general_functions.R")

season.names<-c("winter","spring","summer","fall")
```


### Load seasonal data, and look at number of records, sites, and years
```{r load seasonal data and look at counts}
load(file=file.path(model_data_dir,"flow_timeseries","dseasonal.Rdata"))


dseasonal<-subset(dseasonal,!is.na(HydrologicGroupAB) & !is.na(flow) & !is.na(rain) & !is.na(precip.e.lag2))


# length(unique(dseasonal$site_no))
# #238 unique gages
# 
length(unique(dseasonal$site_no[dseasonal$large_barriers==0]))
length(unique(dseasonal$site_no[dseasonal$TNC_DamCount==0]))
length(unique(dseasonal$site_no[dseasonal$OnChannelWaterSqKM<.5]))
length(unique(dseasonal$site_no[dseasonal$OnChannelWaterSqKM<.5 & dseasonal$large_barriers==0]))
length(unique(dseasonal$site_no[dseasonal$OnChannelWaterSqKM<.5 & dseasonal$TNC_DamCount==0]))
#120 with no LARGE barriers
#97 with no  barriers 


# 7870 total records
nrow(dseasonal)

# 4140 records from sites w/ no LARGE barriers
nrow(dseasonal[dseasonal$large_barriers==0,])
nrow(dseasonal[dseasonal$OnChannelWaterSqKM<.5,])
# 3879 records from sites w/ no LARGE barriers and <.5 sq km impoundments
nrow(dseasonal[dseasonal$OnChannelWaterSqKM<.5 & dseasonal$large_barriers==0,])

# 3283 records from sites w/ no barriers
nrow(dseasonal[dseasonal$TNC_DamCount==0,])
# 3114 records from sites w/ no barriers and <.5 sq km impoundments
nrow(dseasonal[dseasonal$OnChannelWaterSqKM<.5 & dseasonal$TNC_DamCount==0,])


#subset of unregulated gages
#   defined as NO dams from TNC barrier inventory
#   AND upstream on-channel open waterbodies < 0.5 sq km
d.unreg<-subset(dseasonal,TNC_DamCount==0 & dseasonal$OnChannelWaterSqKM<.5)

```

### Create calibration and validation sets
```{r calibration data set}
# View(d.unreg[,c(1:9,58,59,21:23)])

#choose validation set
#select some gages to leave out for validations
#    AND select some years, leave out records from all gages for those year for validation
#    so that combined validation set #records is 10-15% of total records


#choose 8 sites w/ more than 4 yrs records
set.seed(933550175)
val.gages<-sample(unique(d.unreg$site_no[d.unreg$qseasonal>12]),size=8,replace=F)  
# val.gages

#choose 5 years
set.seed(993889335)
val.year<-sample(unique(d.unreg$year),5,replace=F)
sort(val.year)

#check number of records chosen as part of validation gages and validation years
nrow(d.unreg) #3114
nrow(d.unreg[d.unreg$qseasonal>12,]) #2870
nrow(subset(d.unreg,site_no %in% val.gages)) #291
nrow(subset(d.unreg,year(as.Date(date)) %in% val.year)) #212
nrow(subset(d.unreg, site_no %in% val.gages & year(as.Date(date)) %in% val.year )) #22

#subset and count size of calib and valid data sets
d.calib<-subset(d.unreg,!(site_no %in% val.gages | year(as.Date(date)) %in% val.year))
nrow(d.calib) #2633
d.valid<-subset(d.unreg, site_no %in% val.gages | year(as.Date(date)) %in% val.year)
nrow(d.valid) #481   
nrow(d.valid)/nrow(d.unreg)*100 #15.44637    approx 15% of total records

#save calibration and validation data
setwd(file.path(model_dir,"c_seasonal_means/calib_data"))
save(d.calib,d.valid,file="calib_valid.Rdata")
```

### Create mixed model
```{r fixed model}

m.fixed<-lm(log(flow) ~ 
               log(drain_area_va) +
               log(non.zero(DrainageClass)) + 
                    log(non.zero(PercentSandy))+
#                     log(non.zero(HydrologicGroupAB))+ log(non.zero(SurficialCoarseC)) + 
               log(non.zero(Forest)) + 
               log(non.zero(Agriculture)) +     
#                     log(non.zero(Impervious))+
               log(non.zero(OffChannelWaterSqKM)) + 
#                     log(non.zero(OffChannelWetlandSqKM))+
#                     log(non.zero(OffChannelWaterSqKM+OffChannelWetlandSqKM)) + 
#                log(non.zero(BasinSlopePCNT))+ 
                    log(non.zero(ReachSlopePCNT)) +
               precip.e +  precip.e.lag1 + 
               precip.e.lag2 +
               1,
          na.action=NULL, #to ensure we attached fitted values w/ correct site_no
         data=d.calib) 

#create model for each season
m.fixed.season<-list()
# for (i in season.names) {
#      temp<-update(m.fixed, subset=season==i)
#      m.fixed.season[[i]]<-temp
# }
# looping didn't work!  model would change when called later on depending on *current* value of i... oh dear

m.fixed.season[["winter"]]<-update(m.fixed, subset=season=="winter")
m.fixed.season$winter<-update(m.fixed.season$winter,formula=.~.
                              -log(non.zero(DrainageClass))
                              -log(non.zero(PercentSandy)) )


m.fixed.season[["spring"]]<-update(m.fixed, subset=season=="spring")
m.fixed.season$spring<-update(m.fixed.season$spring,formula=.~.
                              -log(non.zero(DrainageClass))
                              -log(non.zero(ReachSlopePCNT))
                              )
                              

m.fixed.season[["summer"]]<-update(m.fixed, subset=season=="summer")
m.fixed.season$summer<-update(m.fixed.season$summer,formula=.~.
                              - log(non.zero(PercentSandy)) 
#                               -log(non.zero(ReachSlopePCNT)) 
                              -log(non.zero(OffChannelWaterSqKM)) 
                              +log(non.zero(Impervious))
                              -log(non.zero(Forest))
                              )


m.fixed.season[["fall"]]<-update(m.fixed, subset=season=="fall")
m.fixed.season$fall<-update(m.fixed.season$fall,formula=.~.
                            -log(non.zero(DrainageClass))
                            -log(non.zero(ReachSlopePCNT))   
                            )

rm(m.fixed)



# summaries of fixed models
for (i in season.names) {
     print(paste("#####",i,"#####"))
     print(summary(m.fixed.season[[i]]))
}

# # all DID...have adj-r-sqr around 0.8976 with old calibration set, but that set was too small....
i<-"winter" #0.827     0.8321 v 0.8249
i<-"spring" #0.8776     0.8929 v .8697
i<-"summer" #0.7356     0.7553 v 0.715
i<-"fall" #0.7833       0.7985 v 0.7827


### next addition: add avg temp, and total number of days <0C
#   might help account for runoff when ground is frozen in winter,
#   and account for ET in summer and fall (when r-sqr values are lower) 
```

```{r stepwise trial, echo=FALSE, eval=FALSE}
# str(d.valid)
# boo<-d.calib[,c("site_no","season","date","flow",
#                 "precip.e","precip.e.lag1","precip.e.lag2","precip.e.lag3",
#                 "da_sqkm","Forest", "Herbacious","Agriculture","Developed",
#                 "DevelopedNotOpen","Impervious",
#                 "DrainageClass","HydrologicGroupAB","HydrologicGroupCD",
#                 "SurficialCoarseC","PercentSandy","ReachElevationM",
#                 "BasinElevationM","ReachSlopePCNT","BasinSlopePCNT",
#                 "OffChannelWaterSqKM","OffChannelWetlandSqKM")]
# boo[,c("da_sqkm","Forest", "Herbacious","Agriculture","Developed",
#                 "DevelopedNotOpen","Impervious",
#                 "DrainageClass","HydrologicGroupAB","HydrologicGroupCD",
#                 "SurficialCoarseC","PercentSandy","ReachElevationM",
#                 "BasinElevationM","ReachSlopePCNT","BasinSlopePCNT",
#                 "OffChannelWaterSqKM","OffChannelWetlandSqKM")]<-
#      apply(boo[,c("da_sqkm","Forest", "Herbacious","Agriculture","Developed",
#                 "DevelopedNotOpen","Impervious",
#                 "DrainageClass","HydrologicGroupAB","HydrologicGroupCD",
#                 "SurficialCoarseC","PercentSandy","ReachElevationM",
#                 "BasinElevationM","ReachSlopePCNT","BasinSlopePCNT",
#                 "OffChannelWaterSqKM","OffChannelWetlandSqKM")],
#            MARGIN=c(1,2),FUN=function(x) log(non.zero(x)))
# 
# 
# i<-"spring" #.8929 v .8697
# i<-"summer" #0.7553 v 0.715
# i<-"fall" #0.7985 v 0.7827
# i<-"winter" #0.8321 v 0.8249
# print(i)
# boo2<-lm(log(flow)~.,data=subset(boo,season==i)[,-c(1:3)])
# boo3<-step(boo2)
# summary(boo3)
# 
# print(summary(m.fixed.season[[i]]))
# 

```

### Create mixed model: random effects for year and site
```{r mixed model- random year and site }


m.mixed<-lmer(log(flow) ~ 
               log(drain_area_va) +
               log(non.zero(DrainageClass)) + 
               log(non.zero(PercentSandy))+
#                     log(non.zero(HydrologicGroupAB))+ log(non.zero(SurficialCoarseC)) + 
               log(non.zero(Forest)) + 
               log(non.zero(Agriculture)) +     
#                     log(non.zero(Impervious))+
               log(non.zero(OffChannelWaterSqKM)) + 
#                     log(non.zero(OffChannelWetlandSqKM))+
#                     log(non.zero(OffChannelWaterSqKM+OffChannelWetlandSqKM)) + 
#                log(non.zero(BasinSlopePCNT))+ 
               log(non.zero(ReachSlopePCNT)) +
               precip.e +  precip.e.lag1 + 
               precip.e.lag2 +
               (1|date)+(1|site_no),
          na.action=NULL, #to ensure we attached fitted values w/ correct site_no
          data=d.calib)

m.mixed.season<-list()

m.mixed.season[["winter"]]<-update(m.mixed, subset=season=="winter")
m.mixed.season$winter<-update(m.mixed.season$winter,formula=.~.
                              -log(non.zero(DrainageClass))
                              -log(non.zero(PercentSandy)) )


m.mixed.season[["spring"]]<-update(m.mixed, subset=season=="spring")
m.mixed.season$spring<-update(m.mixed.season$spring,formula=.~.
                              -log(non.zero(DrainageClass))
                              -log(non.zero(ReachSlopePCNT))
                              )
                              

m.mixed.season[["summer"]]<-update(m.mixed, subset=season=="summer")
m.mixed.season$summer<-update(m.mixed.season$summer,formula=.~.
                              - log(non.zero(PercentSandy)) 
                              -log(non.zero(OffChannelWaterSqKM)) 
                              +log(non.zero(Impervious))
                              -log(non.zero(Forest))
                              )


m.mixed.season[["fall"]]<-update(m.mixed, subset=season=="fall")
m.mixed.season$fall<-update(m.mixed.season$fall,formula=.~.
                            -log(non.zero(DrainageClass))
                            -log(non.zero(ReachSlopePCNT))   
                            )


rm(m.mixed)



```

### Calculate goodness of fit for calibration data set
```{r goodness of fit for CALIBRATION, warning=FALSE}

#create pred df
pred.calib<-d.calib
pred.calib$obs<-log(pred.calib$flow)
pred.calib$pred.fixed<-NA
pred.calib$pred.mixed<-NA

rm(goodness.fixed.calib,goodness.mixed.calib)

#predict and calc goodness of fit metrics of fixed and mixed models, calibration set
for (i in season.names) {
     pred.calib[pred.calib$season==i,"pred.fixed"]<-
          predict(m.fixed.season[[i]],
                  newdata=pred.calib[pred.calib$season==i,])
     pred.calib[pred.calib$season==i,"pred.mixed"]<-
          predict(m.mixed.season[[i]],
                  newdata=pred.calib[pred.calib$season==i,],
                  allow.new.levels=T)
     temp<-goodness(
               pred.calib[pred.calib$season==i,c("obs","pred.fixed")]
               )
     if(!exists("goodness.fixed.calib")){
          goodness.fixed.calib<-temp
          goodness.fixed.calib$season<-i
          goodness.fixed.calib<-goodness.fixed.calib[,c(ncol(goodness.fixed.calib),1:(ncol(goodness.fixed.calib)-1))]
     }
     else
          goodness.fixed.calib[nrow(goodness.fixed.calib)+1,]<-c(i,temp)
     
     temp<-goodness(
               pred.calib[pred.calib$season==i,c("obs","pred.mixed")]
               )
     if(!exists("goodness.mixed.calib")){
          goodness.mixed.calib<-temp     
          goodness.mixed.calib$season<-i
          goodness.mixed.calib<-goodness.mixed.calib[,c(ncol(goodness.mixed.calib),1:(ncol(goodness.mixed.calib)-1))]
     }
     else
          goodness.mixed.calib[nrow(goodness.mixed.calib)+1,]<-c(i,temp)
}

```

### Calculate goodness of fit for validation data set
```{r goodness of fit and predictions for VALIDATION, warning=FALSE}
#create pred df
pred.valid<-d.valid
pred.valid$obs<-log(pred.valid$flow)
pred.valid$pred.fixed<-NA
pred.valid$pred.mixed<-NA

rm(goodness.fixed.valid,goodness.mixed.valid)

#predict and calc goodness of fit metrics of fixed and mixed models, calibration set
for (i in season.names) {
     pred.valid[pred.valid$season==i,"pred.fixed"]<-
          predict(m.fixed.season[[i]],
                  newdata=pred.valid[pred.valid$season==i,])
     pred.valid[pred.valid$season==i,"pred.mixed"]<-
          predict(m.mixed.season[[i]],
                  newdata=pred.valid[pred.valid$season==i,],
                  allow.new.levels=T)
     temp<-goodness(
               pred.valid[pred.valid$season==i,c("obs","pred.fixed")]
               )
     if(!exists("goodness.fixed.valid")){
          goodness.fixed.valid<-temp
          goodness.fixed.valid$season<-i
          goodness.fixed.valid<-goodness.fixed.valid[,c(ncol(goodness.fixed.valid),1:(ncol(goodness.fixed.valid)-1))]
     }
     else
          goodness.fixed.valid[nrow(goodness.fixed.valid)+1,]<-c(i,temp)
     
     temp<-goodness(
               pred.valid[pred.valid$season==i,c("obs","pred.mixed")]
               )
     if(!exists("goodness.mixed.valid")){
          goodness.mixed.valid<-temp     
          goodness.mixed.valid$season<-i
          goodness.mixed.valid<-goodness.mixed.valid[,c(ncol(goodness.mixed.valid),1:(ncol(goodness.mixed.valid)-1))]
     }
     else
          goodness.mixed.valid[nrow(goodness.mixed.valid)+1,]<-c(i,temp)
}
```

### Goodness of fit statistics
```{r view goodness of fit info}

goodness.fixed.calib
goodness.mixed.calib

goodness.fixed.valid
goodness.mixed.valid
```

### Melt data frames for ggplot
```{r melt data frames for ggplot}
#create melted data frames for ggplot

#for correlation plots
dd.corr.calib<-melt(
     pred.calib[,c("site_no","season","year","obs","pred.fixed","pred.mixed")],
     id.vars=c("site_no","season","year","obs"))
dd.corr.calib$season<-factor(capitalize(dd.corr.calib$season),levels=c("Winter","Spring","Summer","Fall"))
dd.corr.valid<-melt(
     pred.valid[,c("site_no","season","year","obs","pred.fixed","pred.mixed")],
     id.vars=c("site_no","season","year","obs"))
dd.corr.valid$season<-factor(capitalize(dd.corr.valid$season),levels=c("Winter","Spring","Summer","Fall"))



#for hydrographs and other plots
dd.calib<-pred.calib[,c("site_no","FEATUREID","date","year","season","qseasonal","da_sqkm",
          "OnChannelWaterSqKM","OnChannelWetlandSqKM","large_barriers","small_barriers",
          "precip.e","precip.e.lag1","precip.e.lag2","flow",
          "obs","pred.fixed","pred.mixed" )]
dd.calib<-melt(dd.calib,id.vars=c("site_no","FEATUREID","date","year","season",
                                  "qseasonal","da_sqkm","OnChannelWaterSqKM",
                                  "OnChannelWetlandSqKM","large_barriers","small_barriers"))

dd.valid<-pred.valid[,c("site_no","FEATUREID","date","year","season","qseasonal","da_sqkm",
          "OnChannelWaterSqKM","OnChannelWetlandSqKM","large_barriers","small_barriers",
          "precip.e","precip.e.lag1","precip.e.lag2","flow",
          "obs","pred.fixed","pred.mixed" )]
dd.valid<-melt(dd.valid,id.vars=c("site_no","FEATUREID","date","year","season",
                                  "qseasonal","da_sqkm","OnChannelWaterSqKM",
                                  "OnChannelWetlandSqKM","large_barriers","small_barriers"))

```

###View correlation of observed vs predicted values
```{r plots correlation, fig.height=8, fig.width=10}
gg.corr<- ggplot(dd.corr.calib,
               aes(x=obs,y=value,colour=variable))+ 
          geom_abline(a=1,b=0,col="black",lty=2)+     
          theme_bw()+xlab("observed")+ylab("predicted")+
          #scale_colour_hue(l=40) 
          scale_colour_hue(l=40, 
               name="Model Type",
               breaks=c("pred.fixed", "pred.mixed"),
               labels=c("Fixed", "Mixed/Random Effects"))
gg.corr + geom_point() + facet_wrap(~season) + ggtitle("Calibration")
gg.corr %+% dd.corr.valid + geom_point() + facet_wrap(~season) + ggtitle("Validation")

```

### View hydrograph of <i>sample</i> sites from calibration set, and all validation sites 
```{r hydrograph of sample sites from calibration set, and all validation sites, fig.height=9, fig.width=14}

gg.hydrograph<-ggplot(data=subset(dd.valid,
                             variable %in% c("obs","pred.fixed","pred.mixed") & site_no %in% val.gages),
                      aes(x=year,y=value,colour=variable)) +
     theme_bw()+scale_colour_brewer(type="qual",palette="Set1")

#pull sample sites from calibration set
set.seed(94802834)
sample.calib<-sample(x=dd.calib[dd.calib$qseasonal>=20,"site_no"],size=10,replace=F)
#calibration hydrographs
for (i in season.names) {
     print(
     gg.hydrograph %+% subset(dd.calib,site_no %in% sample.calib & variable %in% c("obs","pred.fixed","pred.mixed"))+
          geom_line(subset=.(season==i)) +
          geom_point(subset=.(season==i),cex=2,pch=1)+
          geom_point(subset=.(season==i & year %in% val.year),cex=4,pch=18)+
          facet_grid(site_no~.)+
          ggtitle(paste("Calibration:",capitalize(i)))
     )
}

#validation hydrographs
for (i in season.names) {
     print(
     gg.hydrograph + 
          geom_line(subset=.(season==i)) +
          geom_point(subset=.(season==i),cex=2,pch=1)+
          geom_point(subset=.(season==i & year %in% val.year),cex=4,pch=18)+
          facet_grid(site_no~.)+
          ggtitle(paste("Validation:",capitalize(i)))
     )
}


```

### Trial of plots for validating by year 
```{r error plots, fig.height=9, fig.width=14}
gg.error<-ggplot(data=dd.valid,
                 aes(x=site_no,y=value,col=variable)) +
     geom_boxplot(fill="grey30",col="grey30")+geom_point(pch=15,cex=3)+
     scale_colour_brewer(type="qual",palette="Set1")+
     facet_grid(season~year, scales="free_x", space="free_x")


#create sample of years from calibration data set
set.seed(96536573)
sample.calib.year<-sample(x=dd.calib[,"year"],size=10,replace=F)
#calibration plot
gg.error %+% subset(dd.calib, variable %in% c("obs","pred.mixed") & year %in% sample.calib.year )+ ggtitle("Calibration")

#validation plot
gg.error %+% subset(dd.valid, variable %in% c("obs","pred.mixed") & year %in% val.year )+ ggtitle("Validation")


```























